\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-12 NAR_DRAFT.tex
starting on abstract
abstract

\begin{abstract}
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text. Text. Text. Text. Text. Text. Text. Text. Text. Text.
Text. Text.
\end{abstract}

@@

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector \bkmcomment{push equation onto separate line} $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have

>>>>

\begin{abstract}
The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in resistance, and the value of tumor heterogeneity as a biomarker.
The identification of low frequency mutations from Next Generation Sequencing data remains challenging.
In general the ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}

@@

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots \mathrm{d}$ of base calls and their associated quality scores, where $\mathrm{d}$ is total read depth.
We want to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation}
  \label{eqn:hypothesis}
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
\end{equation}
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
The posterior probability of a given hypothesis, $\mathrm{P}(m,\nu)$, is the product of the likelihood of the data given the hypothesis and the prior probability of the hypothesis. 
Assuming that reads are independent, we have
